# Some Self-Contained Examples

```{r include=FALSE, cache=FALSE}
# LIBRARY CONFIGURATION
library(cmdstanr)
library(posterior)
library(knitr)
  knitr::opts_chunk$set(
    cache = TRUE,
    collapse = TRUE,
    comment = NA,
    dev = "png",
    dpi = 150,
    echo = FALSE,
    fig.align = "center",
    fig.width = 6,
    fig.asp = 0.618,  # 1 / phi
    fig.show = "hold",
    include = TRUE,
    message = FALSE,
    out.width = "70%",
    tidy = FALSE,
    warning = FALSE
  )

# GENERAL R CONFIGURATION

options(digits = 2)
options(htmltools.dir.version = FALSE)


set.seed(1123)
```
Time series competition.  A simple example in which we push through
  all the steps from data processing, through Bayesian modeling and
  inference, to decision analysis.  This example also demonstrates how
  to fit a mixture model in Stan.

A few years ago someone sent us an email about a "Global Climate
Challenge" that he had seen online, and which was introduced as
follows:

> It has often been claimed that alarm about global warming is
  supported by observational evidence. I have argued that there is no
  observational evidence for global-warming alarm: rather, all claims
  of such evidence rely on invalid statistical analyses.

> Some people, though, have asserted that the statistical analyses are
  valid. Those people assert, in particular, that they can determine,
  via statistical analysis, whether global temperatures have been
  increasing more than would be reasonably expected by random natural
  variation. Those people do not present any counter to my argument,
  but they make their assertions anyway.

> In response to that, I am sponsoring a contest: the prize is
  $100,000. In essence, the prize will be awarded to anyone who can
  demonstrate, via statistical analysis, that the increase in global
  temperatures is probably not due to random natural variation.

How to win the money?

> The file `Series1000.txt` contains 1000 time series. Each
  series has length 135: the same as that of the most commonly studied
  series of global temperatures (which span 1880–2014). The 1000
  series were generated as follows. First, 1000 random series were
  obtained (via a trendless statistical model fit for global
  temperatures). Then, some randomly-selected series had a trend added
  to them. Some trends were positive; the others were negative. Each
  individual trend was 1°C/century (in magnitude)—which is greater
  than the trend claimed for global temperatures.

> A prize of $100,000 (one hundred thousand U.S. dollars) will be
  awarded to the first person who submits an entry that correctly
  identifies at least 900 series: which series were generated by a
  trendless process and which were generated by a trending process.

But also this:

> Each entry must be accompanied by a payment of $10.

OK, now it's time to get to work.  We start by downloading and
graphing the data.
```{r}
series <- matrix(scan("Series1000.txt"), nrow=1000, ncol=135, byrow=TRUE)
T <- 135
N <- 1000
par(mar = c(3, 3, 2, 0), tck = -0.01, mgp = c(1.5, 0.5, 0))
plot(c(1,T), range(series), bty = "l", type = "n", 
     xlab = "Time", ylab = "series")
for (n in 1:N) {
  lines(1:T, series[n, ], lwd = .5)
}
```

Aha! The lines are fanning out from a common starting point. We'll fit
a regression to each line and then summarize each line by its average
slope.  This is not necessarily the most efficient way to estimate a slope from correlated data, but we will not be concerned about that, as our purpose here is to demonstrate decision analysis in the context of a fitted model, rather than to fit the model in an optimal way.

```{r}
slope <- rep(NA, N)
se <- rep(NA, N)
for (n in 1:N) {
  data <- series[n, ]
  time <- 1:T
  fit <- lm(data ~ time)
  coefs <- summary(fit)$coefficients
  slope[n] <- 100 * coefs[2, "Estimate"]
  se[n] <- 100 * coefs[2, "Std. Error"]
}
```

We multiplied the slopes (and standard errors) by 100 to put them on a
per-century scale to match the above instructions.

Next we plot the estimated slopes and their standard errors:

```{r}
plot(slope, se, bty = "l", xlab = "Slope", ylab = "SE", pch = 20)
```

OK, not much information in the se's. How about a histogram of the
estimated slopes?

```{r}
hist_breaks <- seq(floor(10 * min(slope)), ceiling(10 * max(slope))) / 10
hist(slope, xlab = "Slope", breaks = hist_breaks)
```

Based on the problem description, I'd expect to see distributions
centered at 0, -1, and 1. It looks like this might be the case.

So let's fit a mixture model. That's easy.  Here's the Stan program:

```{r}
cat(paste(readLines("mixture.stan"), "\n", sep=""), sep="")
```

We now run the program and display the results:

```{r}
y <- slope
K <- 3
mu <- c(0, -1, 1)
data <- list(y = y, K = K, N = N, mu = mu)
mod <- cmdstan_model("mixture.stan")
fit_mix <- mod$sample(data = data, parallel_chains = 4, refresh = 0)
print(fit_mix)
```

Convergence is fine: $\widehat{R}$ is close to 1 for everything.  The
estimated weights of the three mixture components are approximately
0.5, 0.25, 0.25. Given that the problem was made up, I'm guessing the
weights of the underlying data-generation process are exactly 1/2,
1/4, and 1/4. The standard deviation of the slopes within each
component is 0.4, or close to it. We could also try fitting a model
where the standard deviations of the three components differ, but we
won't, partly because the description given with the simulated data
described the change as adding a trend, and partly because the above
histogram doesn't seem to show any varying of the widths of the
mixture components.

OK, now we're getting somewhere. To make predictions, we need to know,
for each series, the probability of it being in each of the three
components. We'll compute these probabilities by adding a generated
quantities block to the Stan program:

```
generated quantities {
  matrix[N, K] p;
  for (n in 1:N) {
    vector[K] p_raw; 
    for (k in 1:K) {
      p_raw[k] = theta[k] * exp(normal_lpdf(y[n] | mu[k], sigma));
    }
    for (k in 1:K) {
      p[n,k] = p_raw[k] / sum(p_raw);
    }
  }
}
```

We then re-fit the model, extract the $p$'s, and average them over the
posterior simulations. This is made simpler using the posterior R package's 
`draws_rvars` format. 

```{r, results=FALSE}
mod <- cmdstan_model("mixture_2.stan")
fit_mix <- mod$sample(data = data, parallel_chains = 4, refresh = 0)
prob_sims <- as_draws_rvars(fit_mix$draws())
prob <- mean(prob_sims$p)
```

We now have a $1000\times 3$ matrix of probabilities. Let's take a
look at the first ten rows

```{r}
round(prob[1:10,], 2)
```

So, the first series is probably drawn from the sloping-upward model;
the second might be from the null model or it might be from the
sloping-downward model; the third, fourth, fifth, sixth, seventh, and
eighth are probably from the null model; the ninth is probably from
the sloping-upward model; and so forth.

We'll now program this: for each of the 1000 series in the dataset,
we'll pick which of the three mixture components has the highest
probability. We'll save the probability and also which component is
being picked.

```{r}
max_prob <- apply(prob, 1, max)
choice <- apply(prob, 1, which.max)
```

And now we can sum this over the 1000 series. We'll compute the number
of series assigned to each of the three choices:

```{r}
print(table(choice))
```

The guesses are not quite in proportion 0.5, 0.25, 0.25. There seem to
be too many guesses of zero slope and not enough of positive and
negative slopes. But that makes sense given the decision problem: we
want to maximize the number of correct guesses so we end up
disproportionately guessing the most common category. That's fine;
it's how it will be.

And we can compute the expected number of correct guesses (based on
the posterior distribution we have here), and the standard deviation
of the number of correct guesses (based on the reasonable
approximation of independence of the 1000 series conditional on the
model). And then we'll print all these numbers:

```{r}
expected_correct <- sum(max_prob)
sd_correct <- sqrt(sum(max_prob * (1 - max_prob)))
round(c(expected_correct, sd_correct), 1)
```

Interesting.  The expected number of correct guesses is 
`r round(expected_correct)`. Not quite the 900 that's needed to win! The standard
error of the number of correct guesses is `r round(sd_correct, 1)`, so 900 is
over 5 standard errors away from our expected number correct. That's bad news!

How bad is it?  We can compute the normal cumulative density function
to get the probability of at least 900 successes; that's

```{r, echo = TRUE}
pnorm(expected_correct, 900, sd_correct)
```
That's a small number; here's its reciprocal:
```{r}
recip <- 1 / pnorm(expected_correct, 900, sd_correct)
recip
```
That's almost a 1-in-200,000 chance of winning the big prize!

But we only have to get _at least_ 900. So we can do the continuity
correction and evaluate the probability of at least 899.5 successes,
which, when inverted, yields:
```{r, echo = TRUE}
1 / pnorm(expected_correct, 899.5, sd_correct)
```
Nope, still no good. For the bet to be worth it, even in the crudest
sense of expected monetary value, the probability of winning would
have to be at least 1 in 10,000. (Recall that the prize is \$100,000
but the cost of entry is \$10.) And that's all conditional on the
designer of the study doing everything exactly as he said, and not
playing with multiple seeds for the random number generator,
etc. After all, he could well have first chosen a seed and generated
the series, then performed something like the above analysis and
checked that the most natural estimate gave only 850 correct or so,
and in the very unlikely event that the natural estimate gave 900 or
close to it, just re-running with a new seed. We have no reason to
think that the creator of this challenge did anything like that; our
point here is only that, even if he did his simulation in a completely
clean way, our odds of winning are about 1 in 200,000---about 1/20th
what we'd need for this to be a fair game.

There is one more thing, though: the data are highly autocorrelated,
so least-squares regression may not be the most efficient way to
estimate these slopes. If we can estimate the slopes more precisely,
we can get more discrimination in our predictions.  Maybe there is a
way to win the game by extracting more information from each series,
but it won't be easy.

You could say that the above all demonstrates the designer's point,
that you can't identify a trend in a time series of this length. But we
don't think it would make sense to draw that conclusion from this
exercise. After all, you can just tweak the parameters in the problem
a bit---or simply set the goal to 800 correct instead of 900---and the
game becomes easy to win. Or, had the game been winnable as initially
set up, you could just up the threshold to 950, and again it would
become essentially impossible to win. Conversely, if the designer of
the challenge had messed up his calculations and set the threshold to
800, and someone had sent in a winning entry, it wouldn't disprove his
claims about climate science, it would just mean he hadn't been
careful enough in setting up his bet.
